Fast solvers for partial differential equations \\ subject to inequalities

### 2023-11-06 talk-dms/slides.tex
sketch out remainder of talk

\begin{frame}{nonlinear multigrid by FAS: \only<1>{V-cycle}\only<2>{F-cycle}}

@@

\begin{columns}
\begin{column}{0.45\textwidth}
\begin{itemize}
\item observed optimality:
\begin{align*}
\text{flops} &= O(N^1) \\
\text{exp evaluations} &= O(N^1) \\
\text{processor time} &= O(N^1)
\end{align*}
\item $N\approx 10^8$ unknowns at $J=11$ levels V-cycle
\item compare $\approx 20\,\mu\,\text{s}/\text{N}$ for Poisson using Firedrake ($P_1$, geometric multigrid)
\end{itemize}
\end{column}
\begin{column}{0.55\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/bratu-time.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item VIs are nonlinear problems, even for linear operators like $-\nabla^2$
\item Newton-multigrid is straightforward in PETSc:
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg -pc_type mg \
    -da_refine J
\end{Verbatim}
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
        \begin{itemize}
        \item[] \texttt{rsls} = reduced space line search
        \end{itemize}
    \item[$\circ$] Newton step equations solved by CG with GMG V-cycles
    \end{itemize}
\item issue: the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] grid-dependent (growing) Newton iterations
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item applying nested iteration (nonlinear F-cycle) resolves this:

\vspace{2mm}
\begin{Verbatim}[xleftmargin=13mm,fontsize=\scriptsize]
./obstacle -da_grid_x 3 -da_grid_y 3 \
    -snes_type vinewtonrsls -ksp_type cg -pc_type mg \
\end{Verbatim}

\vspace{-1.5mm}
\hspace{18.7mm} {\scriptsize \color{FireBrick} \texttt{-snes\_grid\_sequence J}}

\vspace{2mm}
    \begin{itemize}
    \item[$\circ$] grid-independent Newton iterations
    \item[$\circ$] optimal $O(N^1)$ flops and time
    \item<2>[$\circ$] Chapter 12 example in my new book
    \end{itemize}

\vspace{-12mm}
\hfill \includegraphics[width=0.25\textwidth]{../talk-oxford/images/mg-fcycle.png}
\end{itemize}

@@

\vspace{-22mm}
\hspace{40mm} \includegraphics[height=0.45\textheight]{../talk-oxford/images/obstacle-flops-per-n.png} \qquad \includegraphics<2>[width=20mm]{../talk-oxford/images/frontcover.jpg}
\end{frame}

@@

{\small
\begin{tabular}{l|ccccc}
   & \stacktwo{admissible}{iterates} & \stacktwo{mesh-indep.}{rates}
       & \stacktwo{no global}{linearization}
         & \stacktwo{PETSc or}{Firedrake}  \\ \hline
RS NM                      & $\checkmark$ &  & & $\checkmark$ \\
\quad $+$ NI & $\checkmark$ & $\checkmark$ & & $\checkmark$ \\
SS NM             &              & $\checkmark$ & & $\checkmark$ \\
{\color{FireBrick} FASCD}  & $\checkmark$ & {\color{FireBrick} \textbf{?}} & $\checkmark$ & \only<2>{{\Large {\color{FireBrick} $\checkmark$}}} {\Large \strut}
\end{tabular}
}

\bigskip
{\scriptsize
\begin{center}
RS = reduced space, SS = semi-smooth, NM = Newton-multigrid, NI = nested iteration
\end{center}
}

\vspace{5mm}
\begin{itemize}
\item for the non-local fluid layer VI problem we need all 4 checked
\item we are trying-out a new algorithm,

{\color{FireBrick} FASCD = full approximation scheme constraint decomposition}

    \begin{itemize}
    \item[$\circ$] Firedrake implementation \only<2>{\dots as of yesterday}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Tai's (2003) \emph{constraint decomposition} (CD) for VIs follows the subspace decomposition idea (Xu 1992)
\item suppose $\mathcal{K} \subset \mathcal{V}$ is a closed and convex admissible subset
\item for a subspace decomposition $\mathcal{V} = \sum_i \mathcal{V}_i$, write the admissible subset as a sum
    $$\mathcal{K} = \sum_i \mathcal{K}_i$$
where $\mathcal{K}^i \subset \mathcal{V}^i$, with projections $\Pi_i : \mathcal{K} \to \mathcal{K}_i$
\item CD additive and multiplicative iterations exist for $VI(F,\ell,\mathcal{K})$

\begin{pseudo}
\pr{cd-add}(u)\text{:} \\+
    for $i \in \{0,\dots,m-1\}$: \\+
        \rm{find} $\hat w_i\in \cK_i$ \rm{so that for all} $v_i\in \cK_i$, \\+
            $\boxed{\ip{F(u - \Pi_i u + \hat w_i)}{v_i-\hat w_i} \ge \ip{\ell}{v_i-\hat w_i}}$ \\--
    $\hat w = \sum_i \hat w_i\in\cK$ \\
    return $w=(1-\alpha) u + \alpha \hat w$
\end{pseudo}
\end{itemize}
\end{frame}

@@

% case of lower obstacle only
\begin{pseudo}
\pr{fascd-vcycle}(\ell^J,\psi^J; w^J)\text{:} \\+
    $\chi^J = \psi^J - w^J {\large \strut}$ \\
    for $j=J$ downto $j=1$ \\+
      $\chi^{j-1} = \maxR \chi^j {\large \strut}$ \\
      $\phi^j = \chi^j - P\chi^{j-1} {\large \strut}$ \\
      $y^j = 0$ \\
      $\text{\pr{smooth}}^{\text{\id{down}}}(\ell^j,\phi^j,w^j;y^j)$ \hspace{8mm} \ct{smoothing in $\mathcal{D}^j$}\\
      $w^{j-1} = \iR(w^j + y^j)$ \\
      $\ell^{j-1} = f^{j-1}(w^{j-1}) + R \left(\ell^j - f^j(w^j+y^j)\right)$ \\-
    $z^0 = 0$ \\
    $\text{\pr{solve}}(\ell^0,\chi^0,w^0;z^0)$ \hspace{10mm} \ct{coarse solve in $\mathcal{U}^0$} \\
    for $j=1$ to $j=J$ \\+
      $z^j = y^{j} + P z^{j-1}$ \\
      $\text{\pr{smooth}}^{\text{\id{up}}}(\ell^j,\chi^j,w^j;z^j)$ \hspace{8mm} \ct{smoothing in $\mathcal{U}^j$} \\-
    $w^J \gets w^J+z^J$
\end{pseudo}
\end{frame}

@@

\begin{frame}{mesh independence}

@@

\begin{frame}{2D shallow ice approximation}

\begin{itemize}
\item preliminary results
\item dome test case in lubrication approximation
    \begin{itemize}
    \item[$\circ$] here $\Phi(s)$ \emph{is} a differential operator
    \item[$\circ$] note $s^{8/3} \in W^{1,4}(\Omega)$ but not in $C^2$
    \end{itemize}
\item FASCD algorithm result
    \begin{itemize}
    \item[$\circ$] Firedrake $P_1$ elements
    \item[$\circ$] strong smoother (\texttt{vinewtonrsls})
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\begin{frame}{nonlinear multigrid by FAS: \only<1>{V-cycle}\only<2>{``FMG'' cycle}}

@@

\begin{columns}
\begin{column}{0.45\textwidth}
\begin{itemize}
\item observed optimality:
\begin{align*}
\text{flops} &= O(N^1) \\
\text{exp evaluations} &= O(N^1) \\
\text{processor time} &= O(N^1)
\end{align*}
\item $N\approx 10^8$ unknowns for $12$-level V-cycle
\item compare $\approx 20\,\mu\,\text{s}/\text{N}$ for Poisson using Firedrake ($P_1$, geometric multigrid)
\end{itemize}
\end{column}
\begin{column}{0.55\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/bratu-time.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{itemize}
\item VIs are nonlinear problems, even for linear operators like $-\nabla^2$
\item most obvious method (?): Newton-multigrid
    \begin{itemize}
    \item[$\circ$] linear solver applies to inactive variables
        \begin{itemize}
        \item[] reduced space line search
        \end{itemize}
    \item[$\circ$] Newton step equations solved by CG with linear multigrid V-cycles
    \end{itemize}
\item issue: the outer Newton iteration must converge on the active set \alert{before} multigrid can provide effective preconditioning
    \begin{itemize}
    \item[$\circ$] consequence: grid-dependent (growing) Newton iterations
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item applying nested iteration helps this:
    \begin{itemize}
    \item[$\circ$] this is effectively an FMG-cycle, an alternative to FASCD
    \item[$\circ$] grid-independent Newton iterations
    \item[$\circ$] optimal $O(N)$ flops and time; Chapter 12 example in my new book
    \end{itemize}
\end{itemize}

@@

\vspace{-22mm}
\hspace{40mm} \includegraphics[width=20mm]{../talk-oxford/images/frontcover.jpg}
\end{frame}

@@

{\small
\begin{tabular}{r|cccc}
   & \stacktwo{admissible}{iterates} & \stacktwo{mesh-indep.}{rates}
       & \stacktwo{no need for}{global linearization}  \\ \hline
NM                         & $\checkmark$ &              & \\
\quad $+$ NI               & $\checkmark$ & $\checkmark$ & \\
{\color{FireBrick} FASCD}  & $\checkmark$ & $\checkmark$ & $\checkmark$ {\Large \strut}
\end{tabular}
}

\bigskip
{\scriptsize
\begin{center}
NM = Newton-multigrid, NI = nested iteration
\end{center}
}

\vspace{5mm}
\begin{itemize}
\item new algorithm,

{\color{FireBrick} FASCD = full approximation scheme constraint decomposition}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Tai's (2003) \emph{constraint decomposition} (CD) for VIs follows the subspace decomposition idea (Xu 1992)
\item suppose $\mathcal{K} \subset \mathcal{V}$ is a closed and convex admissible subset
\item for a subspace decomposition $\mathcal{V} = \sum_i \mathcal{V}_i$, write the admissible subset as a sum
    $$\mathcal{K} = \sum_i \mathcal{K}_i$$
where $\mathcal{K}^i \subset \mathcal{V}^i$, with projections $\Pi_i : \mathcal{K} \to \mathcal{K}_i$
\item CD multiplicative iterations for $VI(F,\ell,\mathcal{K})$

\begin{pseudo}
\pr{cd-mult}
\end{pseudo}
\end{itemize}
\end{frame}

@@

% case of lower obstacle only
\begin{pseudo}
FIXME
\end{pseudo}
\end{frame}

@@

\begin{frame}{classical obstacle problem}

\only<1-2>{\includegraphics[width=0.4\textwidth]{../talk-oxford/images/obstacle65.pdf}}
\hfill
\only<1-2>{\includegraphics[width=0.25\textwidth]{../paper/fixfigs/ball-set.png} \hspace{1mm}}
\only<3>{\includegraphics[width=0.25\textwidth]{../paper/fixfigs/spiral-set.png} \hspace{2mm}}

\begin{itemize}
\item<1> plot $m_J$ versus V-cycle iterations for ball from Table 2 and fit $\log(m_J)$
\item<2> add $m_J$ versus F-cycle iterations for ball from Table 2
\item<3> plot $m_J$ versus V-cycle, F-cycle iterations for spiral from Table 2 and fit $\log(m_J)$
\end{itemize}
\end{frame}

\begin{frame}{advection-diffusion of a concentration}

\mbox{
\includegraphics[width=0.3\textwidth]{../paper/fixfigs/poll2d-zero-set.png}
\includegraphics[width=0.3\textwidth]{../paper/fixfigs/poll2d-one-set.png}
}

\begin{itemize}
\item 2D problem: $-\eps \grad^2 u + \bm{X}\cdot \grad u = \phi$ with $0\le u\le 1$
\item plot $m_J$ versus FMG-cycle iterations from Table 3
\end{itemize}
\end{frame}

@@

\begin{frame}{weak-scaling results: shallow ice approximation}

\hfill \includegraphics[width=0.8\textwidth]{../paper/fixfigs/sialev8scene.png}

\begin{itemize}
\item bar plot $P$ on $x$-axis and time on $y$-axis; label bars with $m_J$ and FMG iterations
\end{itemize}
\end{frame}
